---
title: "Examples of the older lineage formaul: Eq X in text"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Set up code chunk options
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = 'center',
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 300)
# Prevent scientific notation (useful for the rate calculation)
options(scipen = 999999)
pkgload::load_all()
```
### Load in all libraries used for data maipulation and plotting
Load the libary and other libraries

```{r installs, eval=FALSE}
# Install basic data analysis packages
#rm(list = ls())
#install.packages("dplyr")     # Data manipulation
#install.packages("tidyr")     # Data tidying
#install.packages("ggplot2")   # Data visualization
#install.packages("knitr")     # Document generation
library(dplyr)    # For data manipulation
library(tidyr)    # For restructuring data
library(ggplot2)  # For visualization
library(knitr)    # For document generation
# Install the package
library(Branchingdemres)  # For kinship analysis
```

### Import demographic rates
Lets import the period based static demographic rates, in the form of a Leslie matrix: A = U + F.

```{r setup}
U_matrix <- readr::read_rds(here::here("data","u_mat.Rds"))
F_matrix <- readr::read_rds(here::here("data","f_mat.Rds"))
```

### Older sisters; age-specific pmfs

Let's investigate Focal's older sisters. Suppose that Focal is 10 years old. We want to know the probabilities that she has 
zero, one, two, three, four, up to a maximum of eight older sisters of exact age 11. The function call

```{r calculate PMF for Focal's older sisters of age 11 when Focal is aged 10}
pmf_foc_10_os_11 <- os_PMF(y = 10, s1 = 11, U_mat = U_matrix, F_mat = F_matrix, Q = 9)
sum(pmf_foc_10_os_11)
df_foc_10_os_11 <- data.frame(number = seq(0,8),
                              prob = pmf_foc_10_os_11,
                              kin = "Older sisters",
                              kin_age = 11,
                              focal_age = "Focal 10 years")
ggplot(df_foc_10_os_11, aes(x = number, y = prob)) +
  geom_bar(position = "dodge", stat = "identity")  + facet_wrap(~focal_age) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8)) +
  theme(legend.position = "top") + 
  ggtitle("The probabilities that Focal has exactly 0,1...,8 older sisters aged 11, when she is 10")

```


Consider all of the possible ages that Focal's older sister could be. In this case, from 11 years old to 11 + max(mother_reproductive_age). For each possible age of older sisters, we have a probability distribution representing the likeliness that Focal has 0,1,2,... of the kin. Below we take the first three central moments from these age-specific PMFs and plot them

```{r calculate statistics from the PMFs for Focal's older sisters, when Focal is aged 10}
pmf_foc_10_os_full <- data.frame()
for(os_age in 11:50){
  temp_df <- data.frame(number = seq(0,8),
                        prob = os_PMF(y = 10, s1 = os_age, U_mat = U_matrix, F_mat = F_matrix, Q = 9))
  temp_df$age <- rep(os_age, 9)
  pmf_foc_10_os_full <- rbind(pmf_foc_10_os_full, temp_df)
}

age_specific_pmfs_os_foc_10 <- pmf_foc_10_os_full %>%
  group_by(age) %>%
    summarise(X = prob*number,
              X2 = prob*number^2,
              X3 = prob*number^3) %>%
  ungroup() %>% 
  group_by(age) %>%
  summarise(mean = sum(X),
            var = sum(X2) - mean^2,
            m3 = sum(X3) + 2*mean^3 - 3*mean*sum(X2),
            skew = m3/((var^(0.5))^(3))) %>%
  ungroup() %>% mutate(Age_focal = "Focal 10 years old")

legend_title_a <- "Age of Focal"
scale_skew <- 300

age_specific_pmfs_os_foc_10 %>% filter(age < 40) %>%
  ggplot(aes(x = age)) + 
  geom_point(aes( y = mean), color = "blue") + 
  geom_line(aes(y = var^(0.5)), color = "red") +
  geom_line(data = age_specific_pmfs_os_foc_10 %>% dplyr::filter(skew < 90, age < 40), aes(y = skew/scale_skew), 
            linetype = "solid", size = 0.75, color = "grey") +
  scale_y_continuous(sec.axis = sec_axis(~.*scale_skew, name = "Skew (grey)"))+
  ylab("Mean (blue), sd (red)") +
  xlab("Age of older sister") + ggtitle("Focal at age 10")

```



Of course, we can look at how these values change as we condiser different ages of Focal, say at age 60.

```{r calculate statistics from the PMFs for Focal's older sisters, when Focal is aged 60}
pmf_foc_60_os_full <- data.frame()
for(os_age in 61:90){
  temp_df <- data.frame(number = seq(0,8),
                        prob = os_PMF(y = 60, s1 = os_age, U_mat = U_matrix, F_mat = F_matrix, Q = 9))
  temp_df$age <- rep(os_age, 9)
  pmf_foc_60_os_full <- rbind(pmf_foc_60_os_full, temp_df)
}

age_specific_pmfs_os_foc_60 <- pmf_foc_60_os_full %>%
  group_by(age) %>%
    summarise(X = prob*number,
              X2 = prob*number^2,
              X3 = prob*number^3) %>%
  ungroup() %>% 
  group_by(age) %>%
  summarise(mean = sum(X),
            var = sum(X2) - mean^2,
            m3 = sum(X3) + 2*mean^3 - 3*mean*sum(X2),
            skew = m3/((var^(0.5))^(3))) %>%
  ungroup() %>% mutate(Age_focal = "Focal 60 years old")

legend_title_a <- "Age of Focal"
scale_skew <- 300

rbind(age_specific_pmfs_os_foc_10 %>% filter(age < 40),
      age_specific_pmfs_os_foc_60 %>% filter(age < 90))%>%
  ggplot(aes(x = age)) + facet_wrap(~Age_focal, scales = "free_x") +
  geom_point(aes( y = mean), color = "blue") + 
  geom_line(aes(y = var^(0.5)), color = "red") +
  geom_line(data = rbind(age_specific_pmfs_os_foc_10 %>% dplyr::filter(skew < 90, age < 40), 
                         age_specific_pmfs_os_foc_60 %>% dplyr::filter(skew < 90, age < 90)),
            aes(y = skew/scale_skew), 
            linetype = "solid", size = 0.75, color = "grey") +
  scale_y_continuous(sec.axis = sec_axis(~.*scale_skew, name = "Skew (grey)"))+
  ylab("Mean (blue), sd (red)") +
  xlab("Age of older sister") 

```

### Older sisters; the total pmf

The total probabilities of older sisters (i.e., irrespective of their age) is given by the convolution of the age-specific pmfs. Here we compare the total pmfs for Focal at age 10 and age 60
(see Theorem 1 in the MS)
```{r Compare pmf of total to age-specific pmf}
list_to_convolve_f10 <- list()
for(os_age in 11:50){
  temp <- pmf_foc_10_os_full %>% filter(age == os_age)
  temp <- temp$prob
  list_to_convolve_f10[[(1+length(list_to_convolve_f10))]] <- list("prob" = temp, "age_os" = os_age)
}

t_prob_10 <- convoluion_nth(length(list_to_convolve_f10), lapply(1:length(list_to_convolve_f10), function(x){list_to_convolve_f10[[x]][["prob"]]}))

pmd_os_foc_10_totals <- data.frame(number = seq(0,8),
                                   prob = t_prob_10,
                                   kin = "Older sisters",
                                   kin_age = "all possible",
                                   focal_age = "Focal 10 years")

### Compare total 
rbind(df_foc_10_os_11 %>% mutate(method = "Sister aged 11"), pmd_os_foc_10_totals %>% mutate(method = "Any aged sister"))  %>%
  ggplot(aes(x = number, y = prob, color = method, fill = method)) + geom_bar(position = "dodge", stat = "identity") +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8)) +
  theme(legend.position = "top") + 
  ggtitle("Comparing the probability that Focal has and older sister aged 11 and the total probability she has an older siter")

list_to_convolve_f60 <- list()
for(os_age in 61:90){
  temp <- pmf_foc_60_os_full %>% filter(age == os_age)
  temp <- temp$prob
  list_to_convolve_f60[[(1+length(list_to_convolve_f60))]] <- list("prob" = temp, "age_os" = os_age)
}

t_prob_60 <- convoluion_nth(length(list_to_convolve_f60), lapply(1:length(list_to_convolve_f60), function(x){list_to_convolve_f60[[x]][["prob"]]}))

pmd_os_foc_60_totals <- data.frame(number = seq(0,8),
                                   prob = t_prob_60,
                                   kin = "Older sisters",
                                   kin_age = "all possible",
                                   focal_age = "Focal 60 years")

  
rbind(pmd_os_foc_10_totals, pmd_os_foc_60_totals)  %>%
  ggplot(aes(x = number, y = prob, color = focal_age, fill = focal_age)) + geom_bar(position = "dodge", stat = "identity") +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8)) +
  theme(legend.position = "top") + 
  ggtitle("Comparing the probable number of sisters when Focal is 10 and 60")

```
### As a final exploration, let's look at the probable numbers of Focal's older sisters, by each age of Focal

````{r The probable numbers of older sisters over Focal's life course}
############### Older sisters over Focal's life ##################
f_int <- which(mothers_age(U_matrix, F_matrix) > 0)%>%length()
os_df <- lapply(0:100, function(focal_age) {
  os_age <- focal_age + seq_len(f_int)
  os_age <- os_age[which(os_age<=100)]
  if(length(os_age)>0){
    clist <- mapply(function(y) {
      os_PMF(focal_age, y, U_mat = U_matrix, F_mat = F_matrix, Q = 9)
    }, os_age, SIMPLIFY = FALSE)
    prob <- convoluion_nth(length(clist), clist)
    return(list("age focal" = focal_age, "prob sis" = prob))}
  else{return(list("age focal" = focal_age, "prob sis" = as.matrix(c(1, rep(0, 8)))))}
  
})

os_df <- do.call("rbind" , lapply(1:length(os_df), function(x){ data.frame(number = seq(0,8),
                                                  prob = os_df[[x]][["prob sis"]],
                                                  age_focal = os_df[[x]][["age focal"]]) }))
## check sum to one
os_df %>% dplyr::group_by(age_focal) %>%
  dplyr::summarise(check = sum(prob)) %>%
  dplyr::ungroup()

os_df$X <- os_df$number*os_df$prob
os_df$X2 <- os_df$number^2*os_df$prob
os_df$X3 <- os_df$number^3*os_df$prob
os_df_final <- os_df %>%
  dplyr::group_by(age_focal) %>%
  dplyr::summarise(expectation = sum(X),
                   s2 = sum(X2),
                   variance = s2 - expectation^2,
                   s3 = sum(X3),
                   m3 = s3 + 2*expectation^3 - 3*expectation*s2,
                   skew = m3/((variance^(0.5))^(3)),
                   cum_p = cumsum(prob),
                   number = number,
                   prob = prob) %>%
  dplyr::ungroup() %>%
  dplyr::transmute(age_Focal = age_focal,
                   expectation = expectation,
                   variance = variance,
                   skew = skew,
                   kin = "Older Sisters",
                   method = "PMF model",
                   inv_cum_p_95 = ifelse(cum_p >= 0.95 , number, 20 ),
                   inv_cum_p_05 = ifelse(cum_p >= 0.05 , number, 20 ),
                   number = number,
                   prob = prob) %>%
  group_by(age_Focal) %>%
  summarise(number_q95 = min(inv_cum_p_95),
            number_q05 = min(inv_cum_p_05),
            age_Focal = age_Focal,
            expectation = expectation,
            variance = variance,
            skew = skew,
            number = number,
            prob = prob,
            kin = "Older Sisters",
            method = "PMF model")

os_df_final %>%
  ggplot(aes(x = age_Focal, y = prob, color = factor(number))) +
  geom_line(aes(group = factor(number))) +
  theme(legend.position = "top") +
  labs(color = "Number of older sisters",
       y = "Probability",
       x = "Age of Focal"
  )

````
